Machine Learning, Building Vintage and Property Values

### 2019-09-04 text/manuscript_assa.Rmd
Small text changes, Table 1 more condensed

---
title: "Machine Learning, Building Vintage and Property Values"
author: "Thies Lindenthal* (University of Cambridge) & Erik B. Johnson (University of Alabama)"
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: Externalities of architecture, machine learning
fig_caption: yes
header-includes:
   - \usepackage{soul}
   - \usepackage{color}
   - \sethlcolor{yellow}
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{graphicx}
   - \usepackage{grffile}
   - \usepackage{bm}
   - \usepackage{amsmath}
   - \usepackage{setspace}
   - \usepackage{multirow}
   - \doublespacing
   - \usepackage[backend=bibtex, style=authoryear]{biblatex}
fig_width: 6
fig_height: 4
output: 
  pdf_document:
    keep_tex: true
bibliography: references.bib
abstract: "This paper first introduces an algorithm that collects pictures of individual buildings from Google Street View. Second, it trains a deep convolutional neural network (CNN) to classify residential buildings into architectural styles, taking into account spatial dependencies of building vintages. Third, it investigates whether architectural styles influence house prices. For re-sales, the architectural style is a significant determinant of transaction prices while no such effect is found for new buildings. This indicates that any premia are for vintage-related quality characteristics and not for a home's ``beauty''." 
---

# Introduction
\let\svthefootnote\thefootnote
\let\thefootnote\relax\footnote{{*Corresponding author: Thies Lindenthal (htl24@cam.ac.uk). Replication files are available from the author's Github repository \href{https://github.com/thies}{https://github.com/thies}. Paul E. Glade and Lukas Heckmann-Umhau are thanked for exceptional research assistance. Carolin Schmidt, Mike Langen, Colin Lizieri, Franz Fuerst, Stanimira Milcheva, D'Maris Coffman, Norm Miller, Stephen Malpezzi, John Clapp and seminar participants at UCL/Bartlett and at the Homer Hoyt Institute offered excellent feedback and critique on earlier versions of this paper.}}
\addtocounter{footnote}{-1}\let\thefootnote\svthefootnote
If new developments were only pleasing to the eye, so Britain's *Building Better, Building Beautiful Commission*, nimbyism would cease and housing supply could finally reach the levels demanded by a growing and more affluent population [@economistlineofbeauty2018]. The inaugural chair of this new governmental commission boldly suggests that we should *``build, as our Georgian and Victorian forebears built* [...]. *All objections to new building would slip away in the sheer relief of the public''* [@scruton2018]. Even Prince Charles, in similar spirit, put forward ten principles for urban growth and architecture that emphasize tradition and aesthetics [@princecharles2014]. When governments and princes occupy themselves with beauty, economists surely may take a closer look as well. In this paper, we aim to identify the architectural styles of residential buildings using computer vision techniques and to empirically search for any transaction price premia associated with these styles. If aesthetics were as strong a force in the built environment as claimed, we should indeed find a significant effect on property prices.

This paper will assess architectural aesthetics heads on and analyze the visual appearance of residential buildings using street level images. Recent work illustrates how street level imagery and machine learning classification is an efficient and powerful combination for measuring previously unobserved characteristics of the urban environment. @Naik2017 describe how neighborhood demographics may impact the physical appearance of neighborhoods. @Gebru2017 use classified vehicle make and model to predict income, race, education, and voting patterns at the precinct level. @Glaeser2016 predict income in New York City. @Naik2016AER create a neighborhood safety based Streetscore which is shown to be highly correlated with neighborhood population density and household income. @Nadai2016 find that greenery and street facing windows contribute to a positive appearance of safety while @Liu2017 evaluate the quality and upkeep of the built environment along Beijing's streets.

@@

In this paper we suggest a method to collect a large number of images of individual UK buildings from Google Street View, classify the depicted buildings using deep convolutional neural networks, combine the derived information with sales price data and, ultimately, estimate marginal prices for the estimated building characteristics. Our work demonstrates that utilizing pretrained convolutional neural networks to detect complex characteristics such as housing vintages with relatively high accuracy comes at low computational costs and is feasible with only modestly-sized training data. No supercomputers are needed, a modern laptop will do.

@@

The aesthetics of different architectural eras make for an easy cocktail party conversation topic -- and they also influence market prices. Using architectural assessments by human experts, @Buitelaar2017 find a sizable premium of 5\% for new buildings in the Netherlands that refer to traditional styles and a staggering 15\% premium for new buildings that closely follow traditional shapes, facade composition and details. Their study carefully disentangles the architectural style from other unobserved characteristics such as building quality, differences in location or year of construction. These controls are crucial, as earlier work has established that age and vintage variables tend to be highly correlated. @Coulson2008, for instance, suggest a non-parametric estimator and establish a U-shaped age function and distinct price discounts for postwar and contemporary vintages (vis-à-vis more historic vintages). @Francke2017a investigate the depreciation of residential real estate in the Netherlands and decompose land versus structure values singling out the effect of "physical deterioration, functional obsolescence and and vintage effects". They find that buildings from the 1930s carry a strong price premium.

@@

Unfortunately, the traditional approaches chosen by e.g. @Buitelaar2017 or @Ahlfeldt2012 do not scale well. Each observation has to be classified into architectural styles by a human expert, which is time-consuming and costly, imposing an upper limit on the number of observations and the level of detail captured for each observation in any given sample. Google Street View offers a solution to this nexus as it captures the images of almost all buildings in many cities around the world at high level of accuracy and detail. The challenge we address in this paper is to extract and utilize building level information from this ubiquitous sensor.

@@

We download all pictures at the highest resolution offered by Google's Street View API\footnote{\href{https://developers.google.com/maps/documentation/streetview/intro}{https://developers.google.com/maps/documentation/streetview/intro}} (640x640 color pixels). We then obtain 2048-dimensional feature vectors for each picture, using the \emph{Inception-v3} deep convolutional neural network [@Szegedy2015]. \emph{Inception-v3} has been trained for the ImageNet Large Scale Visual Recognition Challenge (ILSVRC)\footnote{\url{http://image-net.org/challenges/LSVRC/}}, which evaluates image classification and object detection algorithms for a wide range of objects. The pretrained classifications would allow us to identify pets, vehicles or people on the pictures -- assessing architectural finesse, however, is beyond the canned classifiers' capabilities.

@@

Two final-year architectural students classified a large sub-sample of approximately 25,000 images from our data set of Cambridge houses. This is a much larger sample than needed in our case. In our case, each category requires less than 250 observations to reach saturated training accuracy levels. We greatly exceed this number for the purpose of this paper so that we can compare the out-of-sample convolutional neural network predictions to the ground truth as assigned by the experts. This allows us to examine the power and size of the assignment tests. In addition having both human and machine classification for a large sample of the data allows for a robustness checks on the machine comparisons.

@@

Two variables are obtained from the distribution of classifications per observation: The ensemble classifier arrives at predictions by majority vote. In addition, a confidence measure is computed as the Herfindahl index of classifications by ensemble member models.\footnote{If, for instance, 20 models classified a building as *Contemporary* but 80 others as *Revival*, the ensemble majority vote is *Revival*. The corresponding Herfindahl score is calculated as $(20/100)^2 + (80/100)^2 = 0.68$.}

@@

Residential real estate transactions are public data in the UK, collected and published by the Land Registry [@LandRegistry2016a]. The records include the date of transaction, price paid, street address, a classification of the property type (flat, detached, semi-detached, or terraced house), the estate type (freehold or leasehold) and an indicator for newly built properties. We select transactions from Cambridge proper which were recorded between January 1995 and October 2018, excluding any leaseholds, flats and properties classified as type "other", and sales with prices below £50,000 or in excess of £2,000,000. Linking recently taken images to sales data from 24 years assumes that houses have not changed their architectural styles -- which seems reasonable. We are not aware of any conversions of just the exterior of buildings. Full redevelopments, did occur and we exclude sales for a given address pre-dating any redevelopment. Table \ref{tab:sumstats} presents summary statistics for the sample.

@@

The Herfindahl scores for classifications vary substantially across ground truth vintages (Table \ref{tab:herftab}), indicating that the models within the ensemble, for instance, agree more often on a building being *Georgian* than finding consensus for other vintages. For *Contemporary* or *Revival* vintages, individual classifications differ most. All off-diagonal elements in the lower panel of Table \ref{tab:herftab} are negative, suggesting that, in general, the ensemble is not really sure when misclassifying images.

@@

Table \ref{tab:imgqualsumstats} presents the mean values for the image quality variables. Table \ref{tab:diffimgchar}, Panel A, compares the mean values for views obstructed by trees or vehicles for correctly (on diagonal) and incorrectly classified images. In most cases, the off-diagonal values are positive, showing that more obstacles in the view of sight correlate with higher odds of misclassifications. Similarly, if more of the image area shows a house (Panel B), misclassifications are reduced (negative differences). This effect is stronger for windows, which appear to give valuable cues about a house's vintage (Panel C). Finally, houses that are not in the center of the building tend to be misclassified more frequently (positive differences in Panel D). In sum, bad quality images lead to misclassifications.

@@

The number of misclassifications decreases decidedly when considering observations with high ensemble confidence only: The majority vote based on the most confident two-thirds of predictions only are more in line with ground truth, as higher recall, precision and $F_1$-scores in Table \ref{tab:confmat}, Panel B, show. The improvement is most evident in the precision rates. For the relatively rare categories of *Georgian* and *Revival* about the same number of buildings classified as *Georgian/Revival* in Panel A actually belong to a different ground truth class -- the precision shoots up to 0.70 for *Georgian* and 0.82 for *Revival* when excluding the not so confident lower third of predictions in Panel B.

@@

Successively, the estimated coefficients from 8 different versions of the hedonic regression specified in Eq. (\ref{eq:hed}) are reported in Tables \ref{tab:hedreg} and \ref{tab:hedregint}. For all models, the hedonic control variables show the expected signs: Negative coefficients for the relative distance to the city center, discounts for terraced homes and semi-detached homes relative to detached houses, positive elasticities for building floor plate and building volumes and a price premium for new buildings compared to second-hand homes, when applicable. Year and neighborhood dummies control for time effects and local amenities but their coefficients are not reported due to space constraints. The combination of location dummies and the distance to the city center measure controls for proximity to the city center \emph{within} each neighborhood.

@@

The first column in Table \ref{tab:hedreg} presents the estimated regression coefficients for ground truth vintages. The base vintage *Contemporary* is more expensive than almost all other vintages, which show negative coefficients that are significantly different from 0. A clear pecking order appears: *Georgian* and *Revival* buildings are the dearest $(+0.04)$,  followed by *Late Victorian/Edwardian* $(+0.01)$, *Contemporary* (base), *Early Victorian* $(-0.13)$,  *Interwar* $(-0.15)$ and *Postwar* $(-0.22)$.

@@

Introducing a variable capturing the most prominent style of buildings on the same street and within 100 m ($\mathbf{VintNeigh}$) does not change the pecking order of building vintages much (Table \ref{tab:hedreg}, Column 7). Neighboring styles are compared to the base case of a *Contemporary* house in a *Contemporary* street. The negative coefficients on dissimilar neighboring styles support a finding by @Lindenthal2017 that shows that a harmonious match of a building's shape with its direct environment leads to a price premium -- or a discount in case of shape mismatches. In relative terms, *Postwar* buildings are the least popular neighbors. The difference between *Revival* and *Contemporary* neighbors is not statistically significant. We fail to find evidence in favor of positive externalities from revival architecture. Everything else equal, buildings within ensembles of *Contemporary* buildings generally do not achieve higher transaction prices than buildings in more historic areas (as Table \ref{tab:comb} will confirm later).

@@

The contributions of this paper are threefold: First, it introduced an algorithm that collects pictures of *individual* buildings from Google Street View. Earlier work has not achieved this level of detail and was, at least in the UK, limited to street sections only. A large-scale application of automatic classification of individual buildings' characteristics using Google Street View has potential not only in the UK. The image collection and classification method can easily be ported to other study areas which have existing Street View data and either LIDAR based building outlines or high resolution satellite images. In a follow-up project, we are working on an improved classification workflow in which a customized object detection model is able to recognize specific building outlines without the need for additional maps or other data.

>>>>

---
title: "Machine Learning, Building Vintage and Property Values"
author: "Thies Lindenthal* (University of Cambridge) & Erik B. Johnson (University of Alabama)"
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: Externalities of architecture, machine learning
fig_caption: yes
header-includes:
   - \usepackage{soul}
   - \usepackage{color}
   - \sethlcolor{yellow}
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{graphicx}
   - \usepackage{grffile}
   - \usepackage{bm}
   - \usepackage{amsmath}
   - \usepackage{setspace}
   - \usepackage{multirow}
   - \doublespacing
   - \usepackage[backend=bibtex, style=authoryear]{biblatex}
fig_width: 6
fig_height: 4
output: 
  pdf_document:
    keep_tex: true
bibliography: references.bib
abstract: "This paper first introduces an algorithm that collects pictures of individual buildings from Google Street View. Second, it trains a deep convolutional neural network (CNN) to classify residential buildings into architectural styles, taking into account spatial dependencies of building vintages. Third, it investigates whether architectural styles influence house prices. For re-sales, the architectural style is a significant determinant of transaction prices while no such effect is found for new buildings. Additionally, through the comparison of expert classification and deep learning methods, we are able to illustrate some of the limitations of machine learning methods and provide guidance on how to detect and overcome these limitations." 
---

# Introduction
\let\svthefootnote\thefootnote
\let\thefootnote\relax\footnote{{*Corresponding author: Thies Lindenthal (htl24@cam.ac.uk). Replication files are available from the author's Github repository \href{https://github.com/thies}{https://github.com/thies}. Paul E. Glade and Lukas Heckmann-Umhau are thanked for exceptional research assistance. Carolin Schmidt, Mike Langen, Colin Lizieri, Franz Fuerst, Stanimira Milcheva, D'Maris Coffman, Norm Miller, Stephen Malpezzi, John Clapp, Pavel Krivenko and participants at the ASSA/AREUEA 2019 annual meeting and seminars at UCL/Bartlett and the Homer Hoyt Institute offered excellent feedback and critique on earlier versions of this paper.}}
\addtocounter{footnote}{-1}\let\thefootnote\svthefootnote
If new developments were only pleasing to the eye, so Britain's *Building Better, Building Beautiful Commission*, nimbyism would cease and housing supply could finally reach the levels demanded by a growing and more affluent population [@economistlineofbeauty2018]. The inaugural chair of this new governmental commission boldly suggests that we should *``build, as our Georgian and Victorian forebears built* [...]. *All objections to new building would slip away in the sheer relief of the public''* [@scruton2018]. Even Prince Charles, in similar spirit, put forward ten principles for urban growth and architecture that emphasize tradition and aesthetics [@princecharles2014]. When governments and princes occupy themselves with beauty, economists surely may take a closer look as well. In this paper, we aim to identify the architectural styles of residential buildings using computer vision techniques and to empirically search for any transaction price premia associated with these styles. If aesthetics were as strong a force in the built environment as claimed, we should indeed find a significant effect on property prices.

This paper assesses the value of architectural aesthetics heads on by classifying the vintage of residential buildings using street level images. Recent work illustrates how street level imagery and machine learning classification is an efficient and powerful combination for measuring previously unobserved characteristics of the urban environment. @Naik2017 describe how neighborhood demographics may impact the physical appearance of neighborhoods. @Gebru2017 use classified vehicle make and model to predict income, race, education, and voting patterns at the precinct level. @Glaeser2016 predict income in New York City. @Naik2016AER create a neighborhood safety based Streetscore which is shown to be highly correlated with neighborhood population density and household income. @Nadai2016 find that greenery and street facing windows contribute to a positive appearance of safety while @Liu2017 evaluate the quality and upkeep of the built environment along Beijing's streets.

@@

In this paper we suggest a method to collect a large number of images of individual UK buildings from Google Street View, classify the depicted buildings using deep convolutional neural networks, combine the derived information with sales price data and, ultimately, estimate marginal prices for the estimated building characteristics. Our work demonstrates that utilizing pretrained convolutional neural networks to detect complex characteristics such as housing vintages with relatively high accuracy comes at low computational costs and is feasible with only modestly-sized training data. Additionally, we illustrate how machine classifications may be sensitive to photo quality by comparing predicted vintages to those made by human experts.

@@

The aesthetics of different architectural vintages\footnote{We use ``style'',``era'' or ``vintage'' interchangeably in this paper.} make for an easy cocktail party conversation topic -- and they may also influence market prices. Using architectural assessments by human experts, @Buitelaar2017 find a sizable premium of 5\% for new buildings in the Netherlands that refer to traditional styles and a staggering 15\% premium for new buildings that closely follow traditional shapes, facade composition and details. Their study carefully disentangles the architectural style from other unobserved characteristics such as building quality, differences in location or year of construction. These controls are crucial, as earlier work has established that age and vintage variables tend to be highly correlated. @Coulson2008, for instance, suggest a non-parametric estimator and establish a U-shaped age function and distinct price discounts for postwar and contemporary vintages (vis-à-vis more historic vintages). @Francke2017a investigate the depreciation of residential real estate in the Netherlands and decompose land versus structure values singling out the effect of "physical deterioration, functional obsolescence and and vintage effects". They find that buildings from the 1930s carry a strong price premium.

@@

Unfortunately, the traditional approaches chosen by e.g. @Buitelaar2017 or @Ahlfeldt2012 do not scale well. Each observation has to be classified into architectural styles by a human expert, which is time-consuming and costly, imposing an upper limit on the number of observations and the level of detail captured for each observation in any given sample. Google Street View offers a solution to this nexus as it captures the images of almost all buildings in many cities around the world at high level of accuracy and detail. The challenge we address in this paper is to extract and utilize building level information from this ubiquitous sensor, using freely available deep learning techniques.

@@

We download all pictures at the highest resolution offered by Google's Street View API\footnote{\href{https://developers.google.com/maps/documentation/streetview/intro}{https://developers.google.com/maps/documentation/streetview/intro}} (640x640 color pixels). We then obtain 2048-dimensional feature vectors for each picture, using the \emph{Inception-v3} deep convolutional neural network [@Szegedy2015]. \emph{Inception-v3} has been trained for the ImageNet Large Scale Visual Recognition Challenge (ILSVRC)\footnote{\url{http://image-net.org/challenges/LSVRC/}}, which evaluates image classification and object detection algorithms for a wide range of objects. The pretrained classifications would allow us to identify pets, vehicles or people on the pictures -- assessing architectural style, however, is beyond the canned classifiers' capabilities.

@@

Two final-year architectural students classified a large sub-sample of approximately 25,000 images from our data set of Cambridge houses. This is a much larger sample than needed in our case. In our case, each category requires less than 250 observations to reach saturated training accuracy levels. We greatly exceed this number for the purpose of this paper so that we can compare the out-of-sample convolutional neural network predictions to the ground truth as assigned by the experts. This allows us to examine the power and size of the assignment tests. In addition having both human and machine classification for a large sample of the data allows for robustness checks on the machine comparisons.

@@

Two variables are obtained from the distribution of classifications per observation: The ensemble classifier arrives at predictions by majority vote. In addition, a confidence measure is computed as the Herfindahl index of classifications by ensemble member models. If, for instance, 20 models classified a building as *Contemporary* but 80 others as *Revival*, the ensemble majority vote is *Revival*. The corresponding Herfindahl score is calculated as $(20/100)^2 + (80/100)^2 = 0.68$.

@@

Residential real estate transactions are public data in the UK, collected and published by the Land Registry [@LandRegistry2016a]. The records include the date of transaction, price paid, street address, a classification of the property type (flat, detached, semi-detached, or terraced house), the estate type (freehold or leasehold) and an indicator for newly built properties. We select transactions from Cambridge proper which were recorded between January 1995 and October 2018, excluding any leaseholds, flats and properties classified as type "other", and sales with prices below £50,000 or in excess of £2,000,000. Linking recently taken images to sales data from 24 years assumes that houses have not changed their architectural styles -- which seems reasonable. We are not aware of any conversions of just the exterior of buildings. Full redevelopments did occur and we exclude sales for a given address pre-dating any redevelopment. Table \ref{tab:sumstats} presents summary statistics for the sample.

@@

The Herfindahl scores for classifications vary substantially across ground truth vintages (Table \ref{tab:herftab}), indicating that the models within the ensemble, for instance, agree more often on a building being *Georgian* than finding consensus for other vintages. For *Contemporary* or *Revival* vintages, individual classifications differ most. All off-diagonal elements in the lower panel of Table \ref{tab:herftab} are negative, suggesting that, in general, the Herfindahl index is a good predictor of misclassified images.

@@

Table \ref{tab:imgqualsumstats} presents the mean values for the image quality variables. Table \ref{tab:diffimgchar}, Panel A, compares the mean values for views obstructed by trees or vehicles for correctly (on diagonal) and incorrectly classified images. In most cases, the off-diagonal values are positive, showing that more obstacles in the view of sight correlate with a higher likelihood of misclassifications. Similarly, if more of the image area shows a house (Panel B), misclassifications are reduced (negative differences). This effect is stronger for windows, which appear to give valuable cues about a house's vintage (Panel C). Finally, houses that are not in the center of the building tend to be misclassified more frequently (positive differences in Panel D). In sum, bad quality images lead to misclassifications.

@@

The number of misclassifications decreases decidedly when considering observations with high ensemble confidence only: The majority vote based on the most confident two-thirds of predictions only are more in line with ground truth, as higher recall, precision and $F_1$-scores in Table \ref{tab:confmat}, Panel B, show. The improvement is most evident in the precision rates. For the relatively rare categories of *Georgian* and *Revival* about the same number of buildings classified as *Georgian/Revival* in Panel A actually belong to a different ground truth class -- the precision shoots up from 0.50 to 0.70 for *Georgian* and from 0.52 to 0.82 for *Revival* when excluding the not so confident lower third of predictions in Panel B.

@@

Successively, the estimated coefficients from 8 different versions of the hedonic regression specified in Eq. (\ref{eq:hed}) are reported in Tables \ref{tab:hedreg} and \ref{tab:hedregint}. For all models, the hedonic control variables show the expected signs: Negative coefficients for the relative distance to the city center, discounts for terraced homes and semi-detached homes relative to detached houses, positive elasticities for building floor plate and building volumes and a price premium for new buildings compared to second-hand homes. Year and neighborhood dummies control for time effects and local amenities but their coefficients are not reported due to space constraints. The combination of location dummies and the distance to the city center measure controls for proximity to the city center \emph{within} each neighborhood.

@@

The first column in Table \ref{tab:hedreg} presents the estimated regression coefficients for ground truth vintages. The base vintage *Contemporary* is more expensive than almost all other vintages, which show negative coefficients that are significantly different from 0. A clear pecking order appears: *Georgian* and *Revival* buildings demand the highest premium $(+0.04)$,  followed by *Late Victorian/Edwardian* $(+0.01)$, *Contemporary* (base), *Early Victorian* $(-0.13)$,  *Interwar* $(-0.15)$ and *Postwar* $(-0.22)$.

@@

Introducing a variable capturing the most prominent style of buildings on the same street and within 100 m ($\mathbf{VintNeigh}$) does not change the pecking order of building vintages much (Table \ref{tab:hedreg}, Column 7). Neighboring styles are compared to the base case of a *Contemporary* house in a *Contemporary* street. The negative coefficients on dissimilar neighboring styles support a finding by @Lindenthal2017 that shows that a harmonious match of a building's shape with its direct environment leads to a price premium -- or a discount in case of shape mismatches. In relative terms, *Postwar* buildings are the least popular neighbors. The difference between *Revival* and *Contemporary* neighbors is not statistically significant. We fail to find evidence in favor of positive externalities from revival architecture. Everything else equal, buildings within ensembles of *Contemporary* generally do not achieve higher transaction prices than buildings in more historic areas (as Table \ref{tab:comb} will confirm later).

@@

The contributions of this paper are threefold: First, it introduces an algorithm that collects pictures of *individual* buildings from Google Street View. Earlier work has not achieved this level of detail and was, at least in the UK, limited to street sections only. A large-scale application of automatic classification of individual buildings' characteristics using Google Street View has potential not only in the UK. The image collection and classification method can easily be ported to other study areas which have existing Street View data and either LIDAR based building outlines or high resolution satellite images. In a follow-up project, we are working on an improved classification workflow in which a customized object detection model is able to recognize specific building outlines without the need for additional maps or other data.
